Motmot: A Distributed Chat Protocol

### 2012-05-04 paper/paper.tex
Introduction

>>>>

We wrote \Motmot due to our frustrations with existing chat protocols. XMPP, the
protocol underlying Google Chat, is a federated XML-based chat protocol. Most
annoyingly, we found that XMPP did not guarantee message delivery---when one
of the users' connection dies, messages sent around the time of the
disconnection may be lost, without any indication in the protocol that this has
happened. Furthermore, XMPP messages are not totally ordered, and different chat
participants can have different chat logs based on when messages were received.

Finally, XMPP was an unsatisfying protocol technologically. While the protocol
itself is federated, all chat messages are still sent to central servers for
relaying. In addition to the obvious single-point-of-failure concerns, there are
also several product-level concerns. In particular, any server which relays your
messages is able to read the entire contents of the chat. Furthermore, your only
guarantee that you are in fact talking to who you think you are comes from the
servers you are talking to.

Other chat protocols besides XMPP share similar properties. Skype, another
popular chat protocol, acts in a peer-to-peer manner.\footnote{Recent rumors
claim that Skype, under Microsoft's kindly direction, has moved some of the
formerly distributed parts of the Skype protocol to a centrally managed server
cluster. In light of such developments, it's unclear just how peer-to-peer
Skype remains.} However, the protocol (and its primary implementation) are
closed-source, so its guarantees cannot be further examined. As firm believers
in sunlight and open protocols, we found Skype also wanting.

\subsection{Design Goals}

We set out implementing \Motmot with several design goals in mind.

\begin{description}
    \item[Multi-party] Chat sessions can occur between any number of
    individuals.
    \item[Reliability] All chat messages should be reliably delivered. Any message
    recorded in the chat history is guaranteed to be delivered to the rest of the
    \Motmot session.
    \item[Total Ordering] All chat participants will receive the same messages
    in exactly the same order.
    \item[Security] All connections in the system will be encrypted with TLS,
    using certificates (and client certificates) to authenticate the other party
    to the communication.
    \item[Deniability] While participants in the chat are confident that the
    messages they receive are sent by the individuals specified, chat
    participants cannot prove the provenance of any message to any third party.
    In other words, after the chat has completed, all participants can fabricate
    any equivalently valid chat history.
    \item[Federation] Anyone can host their own server and issue authentication
    credentials. Any \Motmot user can talk to any other \Motmot user (with an
    identity in any realm) who correctly implements the protocol.
    \item[Presence] Users can maintain buddy lists, and know the connection
    status (online, busy, idle, etc.) of any buddy.
\end{description}
